\dppTitle

### 2014-05-28 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
Working on reviews of dpp-msbayes ms.
Incorporating comments of reviewer 1.

This inference problem is challenging due to the stochastic nature by which
mutations occur in populations and how they are inherited over generations
\cite{Hudson1990,WakeleyCoalescent}.
Thus, a method for estimating historical patterns of divergences across taxa
should explicitly model the stochastic mutational and ancestral processes that
generate and filter the genetic variation we observe in present-day genetic
data.
An appealing approach would be a comparative, Bayesian model-choice method for
inferring the probability of competing divergence histories while integrating
over uncertainty in mutational and ancestral processes via models of nucleotide
substitution and lineage coalescence.

@@

@@

Recently, Oaks et al.\ \cite{Oaks2012,Oaks2014reply} found via simulation that
\msb will often strongly support biogeographically interesting models with a
small number of divergence events shared among taxa, even when divergences were
random over broad timescales.
They suggested this behavior was due to the combination of uniform priors on
parameters causing small marginal likelihoods of richer models and the
U-shaped prior on divergence models.
Hickerson et al.\ \cite{Hickerson2013} suggested the problem is caused by
sampling error, and proposed as a solution an approximate-Bayesian model
averaging approach that samples over empirically informed uniform priors.
However, Oaks et al.\ \cite{Oaks2014reply} evaluated the approach proposed by
Hickerson et al.\ \cite{Hickerson2013} using simulations and found that it did
not mitigate the method's propensity to incorrectly infer clustered
divergences, and often preferred priors that excluded the true values of the
model's parameters.
Here, I describe a new method that successfully mitigates spurious inference
of biogeographically interesting scenarios while avoiding negative side effects
of empirically informed uniform priors.

@@

\subsubsection{Flexibility in parameterizing the model}
In the new implementation, I provide the ability to control the richness of the
model.
For the \myTheta{} parameters, the model can be fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, any model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.

@@

In the original descriptions of the \msb model \cite{Hickerson2006,Huang2011},
this re-ordering is justified by the fact that the expected value of $\pi_b$ is
unrelated to sample size \popSampleSize{1}{} and \popSampleSize{2}{} and thus
exchangeable among pairs.
However, this is incorrect for two reasons.
First, the entire vector of summary statistics \alignmentSS{}{} for each pair
of populations is re-ordered across pairs, which implies that the justification
for re-ordering $\pi_b$ applies to all the statistics within each
\alignmentSS{}{}.
However, the expectations for statistics that estimate gross diversity (e.g.,
$\pi$ and $\theta_W$) are not independent of sample size for structured
populations, and other statistics are not independent of sample size in general
(e.g., $SD(\pi-\theta_W)$).
Second, and more importantly, having the same expectation does not ensure
random variables are exchangeable.
Rather, for variables to be exchangeable their marginal distributions must be
the same (i.e., they must be identically distributed).
None of the summary statistics used by \msb, including $\pi_b$, are identically
distributed across different sample sizes.
In addition to differences in sample size, the simulated alignments and their
summary statistics are not identically distributed because of the taxon- and
locus-specific parameters \hkyModel{}{}, \mutationRateScalarConstant{}{}, and
\ploidyScalar{}{}.

@@

Next, I simulated datasets from three series of models in which the divergence
times of the 22 pairs were random (i.e., no clustering; $\divTimeNum = 22$).
The models comprising each series differ in the variance of the distribution
from which the divergence times are randomly drawn.
When the variance of random divergence times is small, all of the models in
Table~\ref{tabPriors} are expected to struggle to detect this variation and
will often incorrectly estimate highly clustered models of divergence (i.e.,
few divergence events).
The goal is to assess how much temporal variation in random divergence times is
necessary before the behavior of the models of Table~\ref{tabPriors} begins to
improve.
This will determine the timescales over which the models can reliably detect
random variation in divergence times and avoid spurious inference of
biogeographically interesting clustered divergence models.

@@

Overall, the results suggest that the \modelDPP and \modelUniform models are
relatively robust in terms of model-choice accuracy, and when model violations
do cause them to be biased, they tend to under-estimate the probability of the
biogeographically interesting model of a single divergence event.
In contrast, the \modelOld model is very sensitive to model violations, and
strongly over-estimates the probability of the one-divergence model whenever
the model is misspecified.
Furthermore, the results suggest that using exponentially distributed priors on
nuisance parameters rather than uniform priors helps the \modelUshaped model
perform better than \modelOld, but it is still hindered by the U-shaped prior
on divergence models and tends to over-estimate the probability of the
one-divergence model whenever there are violations of the model.

\subsection{Validation analyses: Ordered divergence models}
The results show that the method performs similarily when sampling over ordered
models of divergence
(Figures S\ref{figValidationAccuracyDPPOrderedDPPOrdered} \&
S\ref{figValidationModelChoiceDPPOrderedDPPOrdered}).
This suggests that the method is not adversely affected by the large increase
in the number of possible discrete models (from 22 unordered to 4140 ordered
models) when there are eight pairs of populations.

@@

\section{Discussion}
My results demonstrate that using alternative priors on parameters and
divergence models improved the behavior of the \msb model.
In the new implementation, model-choice estimation is more accurate and shows
greater robustness to model violations
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
The original model is very sensitive to violations and, when present,
strongly over-estimates the probability of one-divergence event shared
across all taxa
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
When more appropriate priors are used for divergence-time and demographic
parameters, and either a Dirichlet-process or uniform prior applied across
divergence models, the model is less sensitive to violations, and, when they do
cause bias, the method is conservative and tends to underestimate the
probability of biogeographically interesting models of temporally clustered
divergences (Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).

@@

I also urge caution when using \dppmsbayes due to the fact that ABC methods
in general are known to be biased for model choice \cite{Robert2011}.
As recommended by Oaks et al.\ \cite{Oaks2012}, any results from the method
should be accompanied by assessments of prior sensitivity and power.
I encourage investigators to view this method as a means of exploring their
data rather than a rigorous means of evaluating hypotheses.

@@

\section{Conclusions}
I introduced a new approximate-Bayesian model-choice method for estimating
shared divergence histories across taxa using DNA sequence data.
The new method, \dppmsbayes, takes a non-parametric approach to the problem by
using a Dirichlet-process prior on the temporal distribution of divergences
across taxa.
The new method shows improved robustness, accuracy, and power compared to the
existing method, \msb.
Compared to \msb, the new approach better estimates posterior uncertainty,
which greatly reduces the chances of incorrectly estimating biogeographically
interesting scenarios of shared divergence events.
This new tool will allow evolutionary biologists to better leverage comparative
genetic data to assess the affects of regional and global biogeographical
processes on biodiversity.

>>>>

This inference problem is challenging due to the stochastic nature by which
mutations occur in populations and how they are inherited over generations
\cite{Hudson1990,WakeleyCoalescent}.
Thus, a method for estimating historical patterns of divergences across taxa
should explicitly model the stochastic mutational and ancestral processes that
generate and filter the genetic variation we observe in present-day genetic
data.
An appealing approach would be a comparative, Bayesian model-choice method for
inferring the probability of competing divergence histories while integrating
over uncertainty in mutational and ancestral processes via models of nucleotide
substitution and lineage coalescence.
\change{
The goal would be to estimate the posterior probability of models ranging from
a single divergence-time parameter (i.e., simultaneous divergence of all
co-distributed taxa) to the fully generalized model in which each taxon
diverged at a unique time.}

@@

\change{
\msb has been used to address biogeographical questions in variety of empirical
systems. Some examples include
(1) whether rising of the Isthmus of Panama caused co-divergence across species
of echinoids co-distributed across the Pacific and Atlantic sides of the ishmus
\cite{Hickerson2006}, 
(2) if an historical seaway across the Baja Peninsula caused co-divergence
across species of squamates and mammals co-distributed both north and south of
the putative seaway \cite{Leache2007},
(3) if species of gall-wasps and their associated parasitoids share divergences
across putative glacial refugia \cite{Stone2012}, and
(4) whether repeated coalescence and fragmentation of the oceanic Islands of
the Philippines caused by Pleistocene sea-level fluctuations caused
diversification of vertebrate taxa distributed across the islands
\cite{Oaks2012}.
Such applications of the method often result in strong posterior support for
co-divergence among some of the taxa investigated (e.g.,
\cite{Barber2010,Carnaval2009,Chan2011,Hickerson2006,Leache2007,Plouviez2009,Stone2012,Voje2009,Oaks2012}).
}

@@

\change{
Recently, Oaks et al.\ \cite{Oaks2012,Oaks2014reply} found via simulation that
\msb will often strongly support models with a small number of divergence
events shared among taxa, even when divergences were random over broad
timescales.}
They suggested this behavior was due to the combination of uniform priors on
parameters causing small marginal likelihoods of richer models and the
U-shaped prior on divergence models.
Hickerson et al.\ \cite{Hickerson2013} suggested the problem is caused by
sampling error, and proposed as a solution an approximate-Bayesian model
averaging approach that samples over empirically informed uniform priors.
However, Oaks et al.\ \cite{Oaks2014reply} evaluated the approach proposed by
Hickerson et al.\ \cite{Hickerson2013} using simulations and found that it did
not mitigate the method's propensity to incorrectly infer clustered
divergences, and often preferred priors that excluded the true values of the
model's parameters.
\change{
Here, I describe a new method that successfully mitigates spurious inference of
co-divergence while avoiding negative side effects of empirically informed
uniform priors.}

@@

\subsubsection{Flexibility in parameterizing the model}
In the new implementation, I provide the ability to control the richness of the
model.
\change{
For the \myTheta{} parameters, by default, the model is fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, if an investigator prefers to reduce the number of parameters, any
model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.}

@@

In the original descriptions of the \msb model \cite{Hickerson2006,Huang2011},
this re-ordering is justified by the fact that the expected value of $\pi_b$ is
unrelated to sample size \popSampleSize{1}{} and \popSampleSize{2}{} and thus
exchangeable among pairs.
\change{
This is incorrect for two reasons.}
First, the entire vector of summary statistics \alignmentSS{}{} for each pair
of populations is re-ordered across pairs, which implies that the justification
for re-ordering $\pi_b$ applies to all the statistics within each
\alignmentSS{}{}.
\change{
However, the expectations for statistics that estimate gross diversity (e.g.,
$\pi$ and $\theta_W$) are not independent of sample size for structured
populations (e.g., the divergent pairs of populations modeled by \msb), and
other statistics are not independent of sample size in general (e.g.,
$SD(\pi-\theta_W)$).}
Second, and more importantly, having the same expectation does not ensure
random variables are exchangeable.
Rather, for variables to be exchangeable their marginal distributions must be
the same (i.e., they must be identically distributed).
None of the summary statistics used by \msb, including $\pi_b$, are identically
distributed across different sample sizes.
In addition to differences in sample size, the simulated alignments and their
summary statistics are not identically distributed because of the taxon- and
locus-specific parameters \hkyModel{}{}, \mutationRateScalarConstant{}{}, and
\ploidyScalar{}{}.

@@

Next, I simulated datasets from three series of models in which the divergence
times of the 22 pairs were random (i.e., no clustering; $\divTimeNum = 22$).
The models comprising each series differ in the variance of the distribution
from which the divergence times are randomly drawn.
When the variance of random divergence times is small, all of the models in
Table~\ref{tabPriors} are expected to struggle to detect this variation and
will often incorrectly estimate highly clustered models of divergence (i.e.,
few divergence events).
The goal is to assess how much temporal variation in random divergence times is
necessary before the behavior of the models of Table~\ref{tabPriors} begins to
improve.
\change{
This will determine the timescales over which the models can reliably detect
random variation in divergence times and avoid spurious inference of clustered
divergence models.}

@@

\change{
Overall, the results suggest that the \modelDPP and \modelUniform models are
relatively robust in terms of model-choice accuracy, and when model violations
do cause them to be biased, they tend to under-estimate the probability of the
model with a single, shared divergence event.}
In contrast, the \modelOld model is very sensitive to model violations, and
strongly over-estimates the probability of the one-divergence model whenever
the model is misspecified.
Furthermore, the results suggest that using exponentially distributed priors on
nuisance parameters rather than uniform priors helps the \modelUshaped model
perform better than \modelOld, but it is still hindered by the U-shaped prior
on divergence models and tends to over-estimate the probability of the
one-divergence model whenever there are violations of the model.

\subsection{Validation analyses: Ordered divergence models}
The results show that the method performs similarily when sampling over ordered
models of divergence
(Figures S\ref{figValidationAccuracyDPPOrderedDPPOrdered} \&
S\ref{figValidationModelChoiceDPPOrderedDPPOrdered}).
This suggests that the method is not adversely affected by the increase
in the number of possible discrete models (from 22 unordered to 4140 ordered
models) when there are eight pairs of populations.
\change{
This is encouraging, because, as discussed above, the sorting of the summary
statistics calculated from sequence alignments is not valid for empirical
datasets in which there is variation in the number of individuals sampled per
population, generation times across taxa, nucleotide substitution parameters,
or the number, length, mutation rate, or ploidy of the loci sampled.
Thus, this sorting should not be performed for any empirical application
of the method.}

@@

\section{Discussion}
My results demonstrate that using alternative priors on parameters and
divergence models improved the behavior of the \msb model.
In the new implementation, model-choice estimation is more accurate and shows
greater robustness to model violations
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
The original model is very sensitive to violations and, when present,
strongly over-estimates the probability of one-divergence event shared
across all taxa
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
\change{
When more appropriate priors are used for divergence-time and demographic
parameters, and either a Dirichlet-process or uniform prior applied across
divergence models, the model is less sensitive to violations, and, when they do
cause bias, the method is conservative and tends to underestimate the
probability of models with temporally clustered divergences
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).}

@@

\change{
I also urge caution when using \dppmsbayes due to the lack of theoretical
validation of Bayesian model choice when the full data are replaced by
insufficient summary statistics, and the recent demonstration that ABC
estimates of model posterior probabilities can be biased when insufficient
summary statistics are employed \cite{Rober2011}.}
As recommended by Oaks et al.\ \cite{Oaks2012}, any results from the method
should be accompanied by assessments of prior sensitivity and power.
I encourage investigators to view this method as a means of exploring their
data rather than a rigorous means of evaluating hypotheses.

@@

\section{Conclusions}
I introduced a new approximate-Bayesian model-choice method for estimating
shared divergence histories across taxa using DNA sequence data.
The new method, \dppmsbayes, takes a non-parametric approach to the problem by
using a Dirichlet-process prior on the temporal distribution of divergences
across taxa.
The new method shows improved robustness, accuracy, and power compared to the
existing method, \msb.
\change{
Compared to \msb, the new approach better estimates posterior uncertainty,
which greatly reduces the chances of incorrectly estimating biogeographical
scenarios of shared divergence events.
This is important, because models of shared divergence events are often of
particular interest to researchers who employ the method.
}
This new tool will allow evolutionary biologists to better leverage comparative
genetic data to assess the affects of regional and global biogeographical
processes on biodiversity.
